\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-24 chapters/conclusion.tex
All review comments

\chapter{Conclusion \& recommendations}
\epigraph{``The mystery of life isn't a problem to solve, but a reality to
experience.''}{S{\o}ren Kierkegaard}
\noindent\emph{The first paragraph of \cref{sec:conclusion/contrib/decoders} was published as
part of \textcite{VanDenKerchove2022}.}

@@

\subsection{Developed \acs{erp} decoders}
\label{sec:conclusion/contrib/decoders}
We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
exploiting prior knowledge about the spatiotemporal nature of the EEG signal
(STBF-struct).
We compared these estimators with the original formulation of the
spatiotemporal (STBF-emp)
beamformer and a state-of-the-art Riemannian Geometry method (XDAWN+RG) in an off-line P3 detection task on
an existing dataset.
Our results show that the structured estimator results in higher
accuracy\todo{qualitative} in
general, and specifically when training data are sparsely available.
Results can be computed faster and with
substantially less memory usage.
Since these algorithms are not paradigm-specific, the conclusions can be
generalized to other ERP-based BCI settings.
These results have been published in~\textcite{VanDenKerchove2022}.

@@

Finally, we developed the \acf{wcble} decoder, specifically designed to address
the challenges posed by P3 latency jitter in covert \ac{vsa} settings.
Latency variability is a well-known issue in \ac{erp} decoding, particularly
when users engage in covert attention, where the timing of the P3 component
fluctuates significantly across trials.
Traditional decoders struggle in these settings due to the inconsistent timing
of brain responses, which can reduce the robustness and accuracy of
\ac{erp} decoding.
The \ac{wcble} decoder mitigates this issue by introducing an iterative process
to estimate and align \ac{erp} latencies across trials.
This is performed in such a way that discriminative power of the \ac{erp}
signal is preserved and enhanced.
The method has been published in~\textcite{VanDenKerchove2024}.
We designed this method to improve decoding where
unseen incoming test data is not yet known.
\Ac{wcble} was first tested on synthetic data to evaluate its
effectiveness under conditions of controlled latency variability and noise.
These results showed that it is robust to higher noise and jitter\ compared to
a non-iterative method in a latency estimation task\todo{qualitative}.
Decoding accuracy was higher in high jitter and noise
settings compared to the non-iterative method and a state-of-the-art
decoder\todo{qualitative}.

@@

Additionally, we gathered data from 7 individuals with \ac{sspgi} with conditions such as
amyotrophic lateral sclerosis (\ac{als}), Friedreich's ataxia (\ac{fa}), and stroke.
These participants exhibited varying degrees of eye motor impairment, such as
involuntary eye movements, ophthalmoplegia, and gaze fixation fatigue.
The data provide invaluable real-world evidence on how gaze-independent
\acp{bci} perform in populations that experience eye motor difficulties.
This dataset allowed us to evaluate the proposed decoding strategies in clinical settings and
highlighted the practical challenges of implementing gaze-independent \acp{bci} in these user
groups.

@@

We evaluated our proposed \ac{wcble} algorithm, its non-iterative
counterpart, and state-of-the-art decoders on the CVSA-ERP dataset as well as on a publicly available
dataset~\cite{Aloise2012} that also contains the overt and covert VSA conditions.
We evaluated the BCI decoding performance in a single-trial classification experiment,
as well as in a target selection experiment reflecting BCI operation.
Performance was significantly different between decoders, but this
result was significantly dependent on the \ac{vsa} condition and the dataset.
While \ac{wcble} was outperformed by the state-of-the-art in overt \ac{vsa}
decoding, it increased covert \ac{vsa} decoding.\todo{qualitative}

@@

To further study the gaze-independent performance of these algorithms, transfer
learning between \ac{vsa} conditions was studied to simulate conditions where an
individual with gaze impairment can end up in different \ac{vsa} settings within
a \ac{bci} operation session due to their lack of proper motor control.
When trained and evaluated on overt VSA data, our proposed \ac{wcble} algorithm
results in a significant decrease in performance compared to the state-of-the-art,
consistent with the within-condition results.\todo{qualitative}
For all other pairs of training and evaluation VSA conditions, however,
\ac{wcble} was equal to or significantly better.\todo{qualitative}

@@

Together, these results show that there is an interest in developing a new class of ERP-BCI
interfaces for users that prefer to choose their own gaze strategy, to
avoid the effort of redirecting their eye gaze to different spatial locations
on the stimulation screen in manners that might be uncomfortable for them.
Coincidentally, advances in gaze-independent decoding also build towards a
solution for the Midas Touch Problem in visual \ac{bci}.
Here, a \ac{bci} user sometimes accidentally selects a target while not intending
to give any input.
Decoding of true intention independent of eye
gaze, with the option of gazing without paying visuospatial attention to a
stimulus, would counteract this.

@@

@@

Finally, we wish to expand our experimentation with individuals with
\ac{sspgi}.
As the work presented in this thesis started during the COVID-19 pandemic,
experimentation was delayed and the envisioned application could not be
validated within the time frame of the doctoral project.
To complete this work, we aim to revisit the participants with an on-line
experiment and comfort assessment to establish the usability of the proposed
interface and compare its \ac{itr} to the state-of-the-art literature.

@@

However, the brain is a complex organ, so neural data is not linear in origin, and
these regularizing assumptions based on prior knowledge are likely only
superficially true.
When optimizing for performance, it will soon become clear that, given a more
flexible model, there are
settings where these assumptions can be broken to better describe the ongoing
interactions.
For example, spatiotemporal beamforming or \ac{lda} with Kronecker-structured covariance
improves performance in low sample sizes but is outperformed by more flexible
methods, such as \ac{lda} using a Kronecker-sum covariance model or \ac{tlda}
when the training sample size increases.
Similarly, \ac{bttda} offers more flexibility.

@@

\subsection{Optimize user experience}
While our work with \ac{bci} users with \ac{sspgi} resulted in interesting
preliminary findings, it is currently impossible to make claims about the experience in long-term
home use of the proposed systems as assistive technology.
This work focused heavily on classification methods and solving problems
arising from gaze-independence through decoding.
A natural metric to then evaluate proposed decoders is \ac{erp} classification
performance.

However, classification performance is only one aspect of what makes a \ac{bci}
desirable as assistive technology.
\Ac{itr} in on-line operation is a better metric, and the ultimate goal
is user satisfaction.
The latter depends on factors such as performance, comfort, and user perception.
Classification performance is a convenient metric because a researcher can perform a single
data collection and iterate on the results offline.
However, this can be too limiting when seeking the most suitable interface
for a specific group of users.

@@

Unfortunately, this is also probably one of the least practical ways to set up
a 4-year research project, such as in the context of a doctoral thesis.
This approach can only work if the proper facilities can be gathered immediately
from the start, such as ethical approval, access to patients, and on-line \acp{bci}
experimentation systems.
It is therefore important that research labs maintain a long-term working
collaboration with patient centers, so that new projects can immediately be
verified with the user population. This also allows interested users to compare
multiple systems and guide development.
The fact that a user has previous \ac{bci} experience is often seen as a confounding
factor in off-line studies.
When the research is in a phase that heavily depends on user
experimentation, this is actually a strength, since the user can help the researcher gain
insight into what works and what does not work for them, and where research efforts
should be focused.

Similarly, there is value in maintaining a working, on-line, in-house \ac{bci}
system that can be iterated upon and adapted if necessary.
Given this system's design and data gathered from earlier use, experiments can
be supported by advance calculations of hypothetical
\ac{itr}, ensuring that on-line or off-line experiments reasonably reflect
real-world operation, and no effort is wasted exploring settings that will not
achieve satisfactory performance.
For example, why study a specific number of repetitions or \ac{isi}, if they
would not be practical to achieve a high \ac{itr}, but could still influence
decoding performance?

This top-down approach is probably not suited for all types of \ac{bci}
research.
Yet, in visual oddball \ac{bci}, where the challenge no longer lies in generating
a proof of concept but in translating the concept into useful technology, this
approach might help avoid prematurely solving problems that may not necessarily
arise further down the road.
A top-down approach could improve the quality of study outcomes by enabling the
reporting of realistic, interpretable metrics,
and by taking into account the human aspect of working \emph{with} instead of
\emph{for} \ac{bci} users.

>>>>

\chapter{Conclusion \& recommendations}
\epigraph{``The mystery of life isn't a problem to solve, but a reality to
experience.''}{ Reverend Mother Gaius Helen Mohiam in \\ \emph{Dune -- Frank
Herbert}\\ adapted from S{\o}ren Kierkegaard}
\noindent\emph{The first paragraph of \cref{sec:conclusion/contrib/decoders} was published as
part of \textcite{VanDenKerchove2022}.}

@@

\subsection{Developed \acs{erp} decoders}
\label{sec:conclusion/contrib/decoders}
We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
exploiting prior knowledge about the spatiotemporal nature of the EEG signal
(STBF-struct).
We compared these estimators with the original formulation of the
spatiotemporal (STBF-emp)
beamformer and a state-of-the-art Riemannian Geometry method (XDAWN+RG) in an off-line P3 detection task on
an existing dataset.
Our results show that the structured estimator results in an accuracy
increase of up to 4 \%. compared to shrinkage regularization.
y when training data are sparsely available.
Results can be computed faster and with
substantially less memory usage.
Since these algorithms are not paradigm-specific, the conclusions can be
generalized to other ERP-based BCI settings.
These results have been published in~\textcite{VanDenKerchove2022}.

@@

Finally, we developed the \acf{wcble} decoder, specifically designed to address
the challenges posed by P3 latency jitter in covert \ac{vsa} settings.
Latency variability is a well-known issue in \ac{erp} decoding, particularly
when users engage in covert attention, where the timing of the P3 component
fluctuates significantly across trials.
Traditional decoders struggle in these settings due to the inconsistent timing
of brain responses, which can reduce the robustness and accuracy of
\ac{erp} decoding.
The \ac{wcble} decoder mitigates this issue by introducing an iterative process
to estimate and align \ac{erp} latencies across trials.
This is performed in such a way that discriminative power of the \ac{erp}
signal is preserved and enhanced.
The method has been published in~\textcite{VanDenKerchove2024}.
We designed this method to improve decoding where
unseen incoming test data is not yet known.
\Ac{wcble} was first tested on synthetic data to evaluate its
effectiveness under conditions of controlled latency variability and noise.
These results showed that it is robust to higher noise and jitter\ compared to
a non-iterative method in a latency estimation task.
Decoding accuracy was higher in high jitter and noise
settings compared to the non-iterative method and a state-of-the-art
decoder.

@@

Additionally, we gathered data from 7 individuals with \ac{sspgi} with conditions such as
\ac{als},\ac{fa}, and stroke.
These participants exhibited varying degrees of eye motor impairment, such as
involuntary eye movements, ophthalmoplegia, and gaze fixation fatigue.
The data provide invaluable real-world evidence on how gaze-independent
\acp{bci} perform in populations that experience eye motor difficulties.
This dataset allowed us to evaluate the proposed decoding strategies in clinical settings and
highlighted the practical challenges of implementing gaze-independent \acp{bci} in these user
groups.

@@

We evaluated our proposed \ac{wcble} algorithm, its non-iterative
counterpart, and state-of-the-art decoders on the CVSA-ERP dataset as well as on a publicly available
dataset~\cite{Aloise2012} that also contains the overt and covert VSA conditions.
We evaluated the BCI decoding performance in a single-trial classification experiment,
as well as in a target selection experiment reflecting BCI operation.
Performance was significantly different between decoders, but this
result was significantly dependent on the \ac{vsa} condition and the dataset.
While \ac{wcble} was slightly outperformed by the state-of-the-art in overt \ac{vsa}
decoding ($94.28<94.74$ and  $85.25<86.60$ \% \ac{rocauc}),
it increased covert \ac{vsa} decoding ($80.84>78.90$ and $73.38>71.11$ \%
\ac{rocauc}).

@@

To further study the gaze-independent performance of these algorithms, transfer
learning between \ac{vsa} conditions was studied to simulate conditions where an
individual with gaze impairment can end up in different \ac{vsa} settings within
a \ac{bci} operation session due to their lack of proper motor control.
When trained and evaluated on overt VSA data, our proposed \ac{wcble} algorithm
results in a small but significant decrease in performance compared to the state-of-the-art
($85.25>86.60$ and $94.74>94.28$ \% \ac{rocauc}), consistent with the within-condition
results.
For all other pairs of training and evaluation VSA conditions, however,
\ac{wcble} was equal to or significantly better with increases exceeding 4 \%
\ac{rocauc}.

@@

Revisiting the hypotheses put forward in
\cref{sec:gaze-independence/objectives}, we conclude that we have proposed
several techniques that did in effect improve visual oddball \ac{erp}-based
\ac{bci} decoding accuracy, and at least one technique that does so specifically
under conditions relevant for gaze-indepent \ac{bci}.
It was not conclusively proven, however, that these results have a meaningful
impact on on-line \ac{bci} assistive technology use by individuals with
\ac{sspgi}.

Together, the obtained results do show that there is an interest in developing
a new class of \ac{erp}-\ac{bci} interfaces for users that prefer to choose their own gaze strategy, to
avoid the effort of redirecting their eye gaze to different spatial locations
on the stimulation screen in manners that might be uncomfortable for them.

@@

Coincidentally, advances in gaze-independent decoding also build towards a
solution for the Midas Touch Problem in visual \ac{bci}.
Here, a \ac{bci} user sometimes accidentally selects a target while not intending
to give any input.
Decoding of true intention independent of eye
gaze, with the option of gazing without paying visuospatial attention to a
stimulus, would counteract this, and be a valuable addition to a \ac{bci}
assistive technology device.

@@

Finally, we wish to expand our experimentation with individuals with
\ac{sspgi}.
As the work presented in this thesis started during the COVID-19 pandemic,
experimentation was delayed and the envisioned application could not be
validated within the time frame of the doctoral project.
To complete this work, we aim to revisit the participants with an on-line
experiment and satisfaction assessment to establish the usability of the proposed
interface according the principles of \ac{ucd} and
compare its \ac{itr} to the state-of-the-art literature.

@@

However, the brain is a complex organ, so neural data is not linear in origin.
These regularizing assumptions based on prior knowledge are likely only
superficially true.
When optimizing for performance, it will soon become clear that, given a more
flexible model, there are
settings where these assumptions can be broken to better describe the ongoing
interactions.
For example, spatiotemporal beamforming or \ac{lda} with Kronecker-structured covariance
improves performance in low sample sizes but is outperformed by more flexible
methods, such as \ac{lda} using a Kronecker-sum covariance model or \ac{tlda}
when the training sample size increases.
Similarly, \ac{bttda} offers more flexibility.

@@

\subsection{Optimize user experience}
While our work with \ac{bci} users with \ac{sspgi} resulted in interesting
preliminary findings, it is currently impossible to make claims about the
experience in long-term home use of the proposed systems as assistive technology.
This work focused heavily on classification methods and solving problems
arising from gaze-independence through decoding.
Since these decoders were evaluated by collecting data and performing tests
later using cross-validation, only off-line performances have been reported.

To gain proper insight in the performance of a system that actually supports
end-users, performance in an on-line experiment should be reported.
The interface, now a collection of white circles, should be adapted
to include meaningful selection elements, like groups of letters or icons and
selection feedback should immediately be presented to the user.
The performance metric then also captures effects of user engagement,
learning and strategy adaptation.
The measured brain activity and the proper execution of the task by the
participant are, after all, partially dependent on their sense of completing a task
that is tangible, entertaining and directly useful for them.
The participant then becomes an active \ac{bci} assistive technology user,
instead of just a near-passive observer in an abstract data collection experiment.

Furthermore, classification performance is only one aspect of what makes a \ac{bci}
desirable as assistive technology.
An interesting approach here is to consider the problem from a
\ac{ucd}~\cite{Standardization2009} perspective.
\Ac{ucd} is a framework for effective optimization and evaluation of usability
in product design, which has successfully been applied to \ac{bci}
development~\cite{Schreuder2013,Kuebler2014,Han2022}.
\Acp{bci} as assistive technologies for users with particular and individual
physical and psychological requirements form a perfect use case for such a framework.

One of the main principles of \ac{ucd} is to address the \emph{entire} user
experience.
As mentioned before, this includes measuring \emph{effectiveness}, in the form
of e.g. selection accuracy.
\Ac{ucd} goes beyond this by also assessing \emph{efficiency} and user
\emph{satisfaction}
Similarly to effectiveness, efficiency can objectively be measured as \ac{itr}
in on-line operation is a better metric.
Satisfaction, on the other hand, is a more subjective quality.

Nevertheless, optimizing user satisfaction is the penultimate goal in
application design.
Satisfaction depends on factors such as performance (in the form of efficiency
and effectiveness), comfort, and user perception.
It must be assessed through subjective questionnaires following realistic, on-line
application use\cite{Kuebler2014}.
Only focusing on evaluating effectiveness through classification performance is
a pragmatic approach, because a researcher can perform a single
data collection and iterate on the results offline.
Yet, this can be too limiting when seeking to design the most usable interface
for a specific group of users.

@@

This approach is tied to another valuable \ac{ucd} principle: encouraging early and active
involvement of users.
Unfortunately, it also might one of the least practical ways to set up
a 4-year research project, such as in the context of a doctoral thesis.
This approach can only work if the proper facilities can be gathered immediately
from the start, such as ethical approval, access to patients, and on-line \acp{bci}
experimentation systems.
It is therefore important that research labs maintain a long-term working
collaboration with patient centers, so that new projects can immediately be
verified with the user population. This also allows interested users to compare
multiple systems and guide development.
The fact that a user has previous \ac{bci} experience is often seen as a confounding
factor in off-line studies.
When the research is in a phase that heavily depends on user
experimentation, this is actually a strength, since the user can help the researcher gain
insight into what works and what does not work for them, and where research efforts
should be focused.

Similarly, there is value in maintaining a working, on-line, in-house \ac{bci}
system that can be iterated upon and adapted if necessary.
Given this system's design and data gathered from earlier use, experiments can
be supported by advance calculations of hypothetical
\ac{itr}, ensuring that on-line or off-line experiments reasonably reflect
real-world operation, and no effort is wasted exploring settings that will not
achieve satisfactory performance.
For example, why study a specific number of repetitions or \ac{isi}, if they
would not be practical to achieve a high \ac{itr}, but could still influence
decoding performance?
Having access to (or implementing) a working, on-line \ac{bci} prototype early
on allows for the implementation of the full \ac{ucd} framework throughout the research
project.
It opens the door for iterative design with interaction between developers and
end-users, yet another \ac{ucd} principle.

This top-down or iterative approach is probably not suited for all types of \ac{bci}
research
If a novel \ac{bci} paradigm is still in an early phase, the focus
should lie on creating limited, fundamental experiments to develop a method to
decode the brain activity of interest and working bottom-up from there to a
proof of concept of an end-user application.
Yet, in visual oddball \ac{bci}, the challenge no longer lies in generating
this proof of concept but in translating it into usable technology.
Here, the top-down approach might help to avoid prematurely solving problems
which may not necessarily arise further down the road.
\Ac{ucd} could improve the quality of study outcomes by enabling the
reporting of realistic, interpretable metrics, and by taking into account the
human aspect of working \emph{with} instead of \emph{for} \ac{bci} users.

### 2024-11-24 chapters/conflict.tex
All review comments

The authors also acknowledge the support by the RITMEA project co-financed by the
European Union with the European Regional Development Fund, the French state,
and the Hauts-de-France Region Council.

>>>>

The authors acknowledge the support by the RITMEA project co-financed by the
European Union with the European Regional Development Fund, the French state,
and the Hauts-de-France Region Council.

### 2024-11-24 chapters/covert_align.tex
All review comments

\Cref{fig:covert-align/block-eval} shows the cross-validated \ac{bci} selection
accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon signed-rank tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure
of \ac{bci} operation intact.
For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.
This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in the
BNCI2014-009 dataset for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
selection accuracy.
No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
\ac{wcble} in CVSA-ERP.
\todo{qualitative, list accuracies for one repetition}

@@

Figure~\ref{fig:covert-align/covert-cross-eval} and Figure~\ref{fig:covert-align/aloise2012-cross-eval} show
the pairwise differences in area under the ROC curve ($\Delta$ROC-AUC)
between the investigated decoders.
In this evaluation, bad epochs were rejected as in \cref{sec:covert-align/results/bloc-acc}.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert-align/covert-cross-eval}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5a.pdf}
	\end{subfigure}

@@

  \caption[Cross-condition classifier performance and estimated jitter.]{
		\subref{fig:covert-align/covert-cross-eval},\subref{fig:covert-align/aloise2012-cross-eval}
    Difference in cross-validated area under the receiver
		operating characteristic curves between Classifier-based Latency Estimation
    (CBLE) and Woody CBLE (WCBLE) across conditions for the CVSA-ERP and
    BNCI2014-009 datasets, respectively.
    A decoder is each	time trained on a visuospatial attention (VSA) condition
    and tested on all VSA conditions.
    WCBLE yields an improvement in most
		non-overt \ac{vsa} settings, indicating it is more invariant to eye gaze than
    CBLE and \ac{tlda}.
		\subref{fig:jitter} Jitter characterized as the interquartile range (IQR)
		of target epochs for different \ac{vsa} conditions. Overt \ac{vsa} exhibits lower
    jitter than other conditions.
    Significance of differences
    was determined with two-sided Wilcoxon signed-rank tests with False Discovery
    Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
    $***=p<0.001$, $****=p<0.0001$).
	}
\end{figure*}
When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
evaluated settings, similar to the within-subject conditions.
\todo{qualitative}
On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
see that performance is on par or greater using \ac{wcble} for most conditions,
except for within-overt \ac{vsa} decoding.\todo{qualitative}

@@

Although it is not immediately clear if \ac{wcble} actively corrects for higher P3 jitter
present in covert and split \ac{vsa} compared to overt \ac{vsa},
we justify our approach in a similar manner to \textcite{Hardiansyah2020} by
observing that the increased discrimination performance of a machine-learning
model accounting for jitter forgoes the need for characterizing the underlying
physiological processes, while still objectively quantifying the presence of
jitter between data classes.
Furthermore, some evidence points toward higher P3 jitter as the main
contributing factor.
While \autoref{fig:erps} shows no visible smearing effect in
the shape of the \acp{erp}, a more quantitative analysis on latencies in
\autoref{fig:jitter} indicates the opposite.
To obtain comparable results across \ac{vsa} conditions, \ac{wcble} was
evaluated per session and trained on all combined \ac{vsa} conditions as in
\cref{sec:covert-align/results/cross}.
Since all conditions have the P3 in common, the estimated latencies can be
interpreted as P3 latencies.
Two-sided Wilcoxon rank-sum tests on jitter, expressed as the interquartile
range (IQR) of the estimated latencies of target trials, revealed that overt
\ac{vsa} exhibited significantly lower jitter than all other conditions for
both datasets.\todo{qualitative}
No other significant differences were found.
$p$-values were corrected for the family-wise error rate using Bonferroni
correction.
Additionally, \textcite{Arico2014} prove that P3 jitter does play a
non-negligible role in covert \ac{vsa} by comparing performance between
overt and covert \ac{vsa}, while including or excluding early VEPs.
This analysis showed that the absence of the N1 and other early VEPs is not the
only factor hampering covert \ac{vsa} decoding performance.
However, the large increase in accuracy for covert \ac{vsa} reported by
\textcite{Arico2014} (60\% to 95\% with a window focusing on only the P3
component) could also be explained by overfitting on artifacts amplified by
aligning, since their method is not evaluated on unseen data, as opposed to ours.

>>>>

\Cref{fig:covert-align/block-eval} shows the cross-validated \ac{bci} selection
accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon signed-rank tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure
of \ac{bci} operation intact.
For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.
This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in the
BNCI2014-009 dataset for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
selection accuracy.
No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
\ac{wcble} in CVSA-ERP.

@@

\Cref{fig:covert-align/covert-cross-eval} and \cref{fig:covert-align/aloise2012-cross-eval} show
the pairwise differences in area under the ROC curve ($\Delta$ROC-AUC)
between the investigated decoders.
In this evaluation, bad epochs were rejected as in \cref{sec:covert-align/results/bloc-acc}.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert-align/covert-cross-eval}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5a.pdf}
	\end{subfigure}

@@

  \caption[Cross-condition classifier performance and estimated jitter.]{
		\subref{fig:covert-align/covert-cross-eval},\subref{fig:covert-align/aloise2012-cross-eval}
    Difference in cross-validated area under the receiver
		operating characteristic curves between Classifier-based Latency Estimation
    (CBLE) and Woody CBLE (WCBLE) across conditions for the CVSA-ERP and
    BNCI2014-009 datasets, respectively.
    A decoder is each	time trained on a visuospatial attention (VSA) condition
    and tested on all VSA conditions.
    WCBLE yields an improvement in most
		non-overt \ac{vsa} settings, indicating it is more invariant to eye gaze than
    CBLE and \ac{tlda}.
		\subref{fig:jitter} Jitter characterized as the interquartile range (IQR)
		of target epochs for different \ac{vsa} conditions. Overt \ac{vsa} exhibits lower
    jitter than other conditions.
    Significance of differences
    was determined with two-sided Wilcoxon signed-rank tests with False Discovery
    Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
    $***=p<0.001$, $****=p<0.0001$).
	}
\end{figure*}
When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
evaluated settings, similar to the within-subject conditions, with the greatest
\ac{rocauc} difference -2\% for training in split $(d=2)$ \ac{vsa} and overt \ac{vsa}.
On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
see that performance is on par or greater using \ac{wcble} for most conditions,
except for within-overt \ac{vsa} decoding, with the greatest \ac{rocauc}
difference +4\% for training in split ($d=2$) \ac{vsa} and testing in overt
\ac{vsa}.

\subsection{Jitter analysis}
\label{sec:covert-align/results/jitter}
Finally, an analysis is performed to quantify the jitter presence in the
different \ac{vsa} conditions.
To obtain comparable results across \ac{vsa} conditions, \ac{wcble} was
evaluated per session and trained on all combined \ac{vsa} conditions as in
\cref{sec:covert-align/results/cross}.
Since all conditions have the P3 in common, the estimated latencies can be
interpreted as P3 latencies.
Results are shown in \cref{fig:jitter}.

Two-sided Wilcoxon rank-sum tests on jitter, expressed as the interquartile
range (IQR) of the estimated latencies of target trials, revealed that overt
\ac{vsa} exhibited significantly lower jitter than all other conditions for
both datasets, with jitter equal to overt: 32 ms and covert:
56 ms ($p<0.001$) for BNCI2014-009, and overt: 33 ms, covert: 77 ms ($p<0.001$),
split ($d=1$): 72 ms ($p=0.006$),
split ($d=2$): 77 ms ($p=0.002$)
and split ($d=3$): 72 ms ($p=0.002$)
for CVSA-ERP.
No other significant differences were found.
$p$-values were corrected for the family-wise error rate using Bonferroni
correction.

@@

Although it is not immediately clear if \ac{wcble} actively corrects for higher P3 jitter
present in covert and split \ac{vsa} compared to overt \ac{vsa},
we justify our approach in a similar manner to \textcite{Hardiansyah2020} by
observing that the increased discrimination performance of a machine-learning
model accounting for jitter forgoes the need for characterizing the underlying
physiological processes, while still objectively quantifying the presence of
jitter between data classes.
Furthermore, some evidence points toward higher P3 jitter as the main
contributing factor.
While \autoref{fig:erps} shows no visible smearing effect in
the shape of the \acp{erp}, the more quantitative analysis on latencies
presented in \cref{sec:covert-align/results/jitter}
indicates the opposite.
Additionally, \textcite{Arico2014} prove that P3 jitter does play a
non-negligible role in covert \ac{vsa} by comparing performance between
overt and covert \ac{vsa}, while including or excluding early VEPs.
This analysis showed that the absence of the N1 and other early VEPs is not the
only factor hampering covert \ac{vsa} decoding performance.
However, the large increase in accuracy for covert \ac{vsa} reported by
\textcite{Arico2014} (60\% to 95\% with a window focusing on only the P3
component) could also be explained by overfitting on artifacts amplified by
aligning, since their method is not evaluated on unseen data, as opposed to ours.
